---
title: "Data Wrangling in R"
author: "Federico Lopez"
date: "November 13, 2015"
output: 
  pdf_document:
    toc: yes
---

# An introduction to manipulating data with `tidyr` and `dplyr`
`tidyr` is designed specifically for data tidying (not general reshaping or aggregating) and works well with `dplyr` data pipelines. Tidy data ensures that values of different variables from the same observation are always paired. In tidy data:

* Each variable forms a column
* Each observation forms a row
* Each type of observational unit forms a table

For further detail see https://github.com/hadley/tidyr and https://ramnathv.github.io/pycon2014-r/explore/tidy.html

`dplyr` provides several functions for manipulating data frames. It is a new iteration of the `plyr` package, which implements the “split-apply-combine” strategy for data analysis. Here we will go over a few examples of data manipulation using `dplyr`.

# Reshaping data
## Gather columns into rows
```{r, warning=FALSE, message=FALSE}
# Install and load tidyr and dplyr
#install.packages("tidyr")
#install.packages("dplyr")
library(tidyr)
library(dplyr)

# Load the Iris data 
# This data frame includes 150 obs. of 5 variables
str(iris)
head(iris)

# Use the tbl_df from dplyr to transform a regular data frame to a tbl object
# "tbl objects only print a few rows and all the columns that fit on one screen"
# (transforming data frames into tbl_dfs is not a requirement to use dplyr)
iris <- tbl_df(iris)
iris

# Summarize tbl data
glimpse(iris)

# Get first observation for each Species in iris data
mini.iris <- iris[c(1, 51, 101), ]
mini.iris

# Gather columns into rows (converting data from wide to long)
# Usage: gather(data, key, value, ...)
# data: A data frame.
# key, value: Names of key and value columns to create in output
# ...: Specification of columns to gather
# -Species is used to drop the Species column and gather the remaining
# columns, that is, Sepal.Length, Sepal.Width, Petal.Length, and Petal.Width
mini.irisl <- gather(mini.iris, key=flower_trait, value=measurement, -Species)
mini.irisl
```

## Spread rows into columns
```{r}
# Spread rows into columns (converting data from long to wide)
mini.irisw <- spread(mini.irisl, flower_trait, measurement)
mini.irisw
```

## Order rows by values
```{r}
# low to high
arrange(iris, Sepal.Length)

# high to low
arrange(iris, desc(Sepal.Length))
```

## Rename the columns of a data frame
```{r}
rename(iris, sp=Species)
```

# Manipulating data (filter, select, mutate, bind etc.)
`dplyr` provides a function for each basic verb of data manipulation:

* `filter()` (and `slice()`)
* `arrange()`
* `select()` (and `rename()`)
* `distinct()`
* `mutate()` (and `transmute()`)
* `summarise()`
* `sample_n()` and `sample_frac()`

The data is always the first argument of the verb functions.

## Subset observations (rows)
### Filter
```{r}
# Extract rows that meet logical criteria
filter(iris, Sepal.Length > 6)
iris[iris$Sepal.Length > 6, ] # Using base R

# Extract rows according to a factor
filter(iris, Species == "setosa")
subset(iris, Species == "setosa") # Using base R

# Provide any number of filtering conditions, 
# which are joined together with & or |
filter(iris, Species == "setosa" & Sepal.Length > 5)
#filter(iris, Species == "setosa" & Sepal.Length > median(Sepal.Length))
# Using base R code
iris[iris$Species == "setosa" | iris$Sepal.Length > median(iris$Sepal.Length), ]
```

### Remove duplicate rows
```{r}
# Remove duplicate rows
# Show duplicated rows: duplicated(iris)
distinct(iris)
```

### Randomly select rows
```{r}
# Randomly select fraction of rows
sample_frac(iris, 0.5, replace=TRUE)

# Randomly select n rows
sample_n(iris, 10, replace=TRUE)

# Select rows by position
slice(iris, 10:15)
```

## Subset variables (columns)
### Select columns by complete or partial name, drop variables
```{r}
# Select columns by name 
select(iris, Sepal.Width, Petal.Length, Species)

# Select columns whose name starts with a character string
select(iris, starts_with("Petal"))

# Select columns whose name ends with a character string
select(iris, ends_with("Length"))

# Select columns whose names are in a group of names
#select(iris, one_of("Species"))

# Drop variables
select(iris, -starts_with("Petal"))
select(iris, -contains("etal"))
select(iris, -Petal.Length, -Petal.Width)
select(iris, -(Sepal.Length:Petal.Length))

# Using base R
# iris$Species <- NULL
# iris[["Species"]] <- NULL
# iris[,"Species"]  <- NULL
# iris[[5]] <- NULL
# iris[,5] <- NULL
# iris <- subset(iris, select=-Species)
```

## Summarize data
```{r, warning=FALSE}
summarise(group_by(iris, Species), mean(Sepal.Length))

# Apply summary function to each column
summarise_each(iris, funs(mean))

iris %>%
  select(-Species) %>%
  summarise_each(funs(mean))
```

## Group data
```{r}
# Group data into rows with the same value of Species
group_by(iris, Species)

# Remove grouping information from data frame
ungroup(iris)
```

## Make new variables
```{r}
# Compute and append one or more new columns
mutate(iris, sepal=Sepal.Length + Sepal.Width)
mutate(iris, logSepal.Length=log10(Sepal.Length))

# Using base R
# iris[["logSepal.Length"]] <- log10(iris$Sepal.Length)
# iris[,"logSepal.Length"] <- log10(iris$Sepal.Length)
# iris$logSepal.Length <- log10(iris$Sepal.Length)

# Drop existing variables
transmute(iris, logSepal.Length=log10(Sepal.Length))

# Compute one or more new columns. Drop original columns
transmute(iris, sepal=Sepal.Length + Sepal.Width) 
```

## Combine data sets
```{r}
a <- data_frame(x1=c("A", "B", "C"), x2=c(1, 2, 3))
a
b <- data_frame(x1=c("A", "B", "D"), x3=c("T", "F", "T"))
b
# Join matching rows from b to a
left_join(a, b, by="x1")

# Join matching rows from a to b
right_join(a, b, by="x1")

# Join data. Retain only rows in both sets
inner_join(a, b, by="x1")

# Join data. Retain all values, all rows
full_join(a, b, by="x1")

# All rows in a that have a match in b
semi_join(a, b, by="x1")

# All rows in a that do not have a match in b
anti_join(a, b, by="x1")
```

### Set operations
These expect the x and y inputs to have the same variables, and treat the observations like sets.
```{r}
y <- data_frame(x1=c("A", "B", "C"), x2=c(1, 2, 3))
z <- data_frame(x1=c("B", "C", "D"), x2=c(2, 3, 4))
str(y)
glimpse(y)
y
z

# Rows that appear in both y and z
intersect(y, z)

# Rows that appear in either or both y and z
union(y, z)

# Rows that appear in y but not z
setdiff(y, z)
```

### Binding
```{r}
# Append z to y as new rows
bind_rows(y, z)

# Append z to y as new columns. Caution: matches rows by position
bind_cols(y, z)
```

For further information on `dplyr`:  
https://cran.r-project.org/web/packages/dplyr/index.html  
http://blog.rstudio.org/2014/01/17/introducing-dplyr/  
http://www.dataschool.io/dplyr-tutorial-for-faster-data-manipulation-in-r/  

# Using `magrittr` to create pipelines
According to its documentation (https://github.com/smbache/magrittr), the `magrittr` package offers a set of operators which promote semantics that will improve your code by:

* structuring sequences of data operations left-to-right (as opposed to from the inside and out),
* avoiding nested function calls,
* minimizing the need for local variables and function definitions, and
* making it easy to add steps anywhere in the sequence of operations.

A simple example:
```{r, warning=FALSE, message=FALSE}
#install.packages("magrittr")
library(magrittr)
iris %>% head(4)
```

The pipe operator `%>%` passes the object on left hand side as first argument of function on righthand side. `dplyr` imports the `%>%` operator from `magrittr`.

```{r}
iris %>%
  filter(Species == "setosa") %>%
  select(Sepal.Length, Sepal.Width) %>%
  head(10)

iris %>%
  group_by(Species) %>%
  summarise(avg=mean(Sepal.Width)) %>%
  arrange(avg)
```
  
There are also functions that do not have a data argument, for which it is useful to expose the variables in the data. This is done with the `%$%` operator:
```{r}
iris %>%
  filter(Sepal.Length > mean(Sepal.Length)) %$%
  cor(Sepal.Length, Sepal.Width)
```

Which country experienced the sharpest 5-year drop in life expectancy?
```{r}
#install.packages("gapminder")
library(gapminder)
gtbl <- tbl_df(gapminder)
glimpse(gtbl)

#worstle <- 
gtbl %>%
  group_by(continent, country) %>%
  select(country, year, continent, lifeExp) %>%
  # lag(): copy with values lagged by 1
  mutate(le.delta=lifeExp - lag(lifeExp)) %>%
  summarize(worstle.delta=min(le.delta, na.rm=TRUE)) %>%
  filter(min_rank(worstle.delta) < 2) %>%
  arrange(worstle.delta)
#worstle
```

What is the correlation between life expectancy and year within each country?
```{r}
# Fit a linear regression within country
library(ggplot2)
str(gapminder)
ggplot(gapminder, aes(x=year, y=lifeExp, colour=continent)) +
  geom_jitter() +
  geom_smooth(lwd=1.5, method="lm")

# Calculate overall correlation between year and life expectancy
(ov.cor <- gapminder %$% 
  cor(year, lifeExp))
#> [1] 0.4356112
# Calculate correlation within each country
(gcor <- gapminder %>%
  group_by(country) %>%
  summarize(correlation=cor(year, lifeExp)))

# The correlation between life expectancy and year is 
# much higher within countries
ggplot(gcor, aes(x=correlation)) +
  geom_density() +
  geom_vline(xintercept=ov.cor, linetype="longdash") +
  geom_text(data=NULL, x=ov.cor, y=10, label=round(ov.cor, 2),
            hjust=-0.1)
```

# Pattern matching and replacement
Some content included here taken from https://github.com/STAT545-UBC

## Position of pattern within the string
* `^`: start of the string.
* `$`: end of the string.
* `\b`: empty string at either edge of a word. Don’t confuse it with ^ $ which marks the edge of a string.
* `\B`: empty string provided it is not at an edge of a word.

## Operators
* `.`: matches any single character, as shown in the first example.
* `[...]`: a character list, matches any one of the characters inside the square brackets. We can also use - inside the brackets to specify a range of characters.
* `[^...]`: an inverted character list, similar to [...], but matches any characters except those inside the square brackets.
* `\`: suppress the special meaning of metacharacters in regular expression, i.e. `$ * + . ? [ ] ^ { } | ( ) \`, similar to its usage in escape sequences. Since `\` itself needs to be escaped in R, we need to escape these metacharacters with double backslash like `\\$`.
* `|`: an “or” operator, matches patterns on either side of the `|`.
* `(...)`: grouping in regular expressions. This allows you to retrieve the bits that matched various parts of your regular expression so you can alter them or use them for building up a new string. Each group can than be refer using `\\N`, with N being the No. of `(...)` used. This is called backreference.

```{r}
# Find matches for a string
?grep
bee.spp <- c("Apis cerana", "Apis koschevnikovi", "Apis mellifera", 
  "Apis nigrocincta", "Bombus atratus", "Bombus dahlbomii", 
  "Bombus fervidus", "Bombus lapidarius", "Bombus ruderatus", 
  "Bombus rupestris")

i <- grep("Bombus", bee.spp)
cat("'Bombus' appears", length(bee.spp[i]), "times")
i
bee.spp[i]

# Replace specified values
gsub(" ", "_", bee.spp)

#grep -oE "^>.*\|" sequence.fasta
#grep -oE "\|.*\|" ants.fasta
```
