---
title: "Multivariate Stats in R"
author: "Mike Herrmann"
date: "Thursday, October 29, 2015"
output: 
  pdf_document:
    toc: yes
---

# Statistical tests that we will be covering
1. ANOVA and LM
2. Principal Component Analysis
3. Discriminant Analysis
4. Non-metric Multidimentional Scaling (NMDS)

```{r eval=FALSE}
# Install required packages
install.packages("MASS")
install.packages("ggplot2")
install.packages("vegan")
install.packages("Ecdat")
install.packages("ellipse")
```
```{r warning=FALSE, message=FALSE}
# Load packages
library(MASS)
library(ggplot2)
library(vegan)
library(Ecdat)
```

# ANOVA, ANCOVA, and LM
First example, basic ANOVA. We will use a diamond dataset in the package Ecdat. It compares the price of diamonds to several measurements.

```{r}
attach(Diamond)
View(Diamond)

#test for effects of transmission type, starting with a simple one-way test

diamond_test1 <- aov(price ~ clarity)
summary(diamond_test1)

#add in more factors, using * between factors will give you a full-factorial design
#A WARNING! R uses a type-I sequencial SS for their analysis. THIS ANALYSIS WILL GIVE DIFFERENT RESULTS BASED ON VARIABLE ORDER FOR NON-ORTHOGONAL DESIGNS! Watch for this if you do not have equal observations in all of your variables!
diamond_test3 <- aov(price ~ colour*clarity*certification)
summary(diamond_test3)
#To look at diagnostic plots, we can use the plot function
plot(diamond_test3)

#to see effect directions, boxplots work well
boxplot(price~clarity)
plot(price~carat)

#For an ANCOVA, replace * with +. This is also useful if you have a random variable, such as with a blocking design, that you want to try to account for.

diamond_ancova<-aov(price ~ clarity + colour)
summary(diamond_ancova)

#linear models work essentially the same way, but instead of "aov", we use "lm"

diamond_lm <- lm(price~carat)
summary(diamond_lm)
plot(diamond_lm)


detach(Diamond)
```

# Principal Component Analysis (PCA)
Principal component analysis- to look at this, we will use a different sample data set that comes with R, the Iris dataset
```{r}
attach(iris)
summary(iris)

par(mfrow=c(1,1))
PC <- princomp(iris[,1:4], scores=TRUE)
summary(PC)
plot(PC, type="lines")

biplot(PC)
biplot(PC, xlabs=Species)
biplot(PC)
#set up a 3D plot
library(rgl)
plot3d(PC$scores[,1:3], col=as.integer(Species))

#We can perform unsupervised clustering using k-means
cl <- kmeans(iris[,1:4],3)
iris$cluster <- as.factor(cl$cluster)
#Now lets check how our clustering performed
table(iris$cluster, Species)
```

# Linear Discriminant Analysis
```{r}
# Check for collinear values
cor <- cor(iris[,1:4])
dissimilarity <- 1-abs(cor)

distance <- as.dist(dissimilarity)

hc <- hclust(distance)

clusterV=cutree(hc, h=0.05)
clusterV

cor(Petal.Length, Petal.Width)
Petal.Width <- NULL
# The LDA
irisLDA <- lda(iris[,1:4], Species, CV=FALSE)
irisCV <- lda(iris[,1:4], Species, CV=TRUE)

plot(irisLDA)
irisLDA$counts
#CV
table(Species, irisCV$class)
pred <- predict(irisLDA)
#Percent explained by each DA
prop <- irisLDA$svd^2/sum(irisLDA$svd^2)
prop

#use ggplot to make a much prettier version of the LDA plot
pred<-data.frame(Species=predict(irisLDA)$class,predict(irisLDA)$x) 
library(ellipse) 
dat_ell <- data.frame() 

for(g in levels(pred$Species)){ 
dat_ell <- rbind(dat_ell, cbind(as.data.frame(with(pred[pred$Species==g,], ellipse(cor(LD1, LD2), scale=c(sd(LD1),sd(LD2)),centre=c(mean(LD1),mean(LD2))))),Species=g))} 

ggplot(pred, aes(x=LD1, y=LD2, col=Species) ) + geom_point( size = 4, aes(color = Species))+theme_bw()+geom_path(data=dat_ell,aes(x=x,y=y,color=Species),size=1,linetype=2) 
```

# Non-metric Multidimentional Scaling (NMDS)
```{r}
set.seed(2)
#set up random species data
community_matrix=matrix(
   sample(1:100,300,replace=T),nrow=10,
   dimnames=list(paste("community",1:10,sep=""),paste("sp",1:30,sep="")))

example_NMDS=metaMDS(community_matrix, # Our community-by-species matrix
                     k=2) # k = number of dimentions

#stressplot - if points remain close to the line, our data retains its original differences despite the dimention reduction
stressplot(example_NMDS)


#plot nmds- communities ("sites") are open circles, species are red crosses
plot(example_NMDS)

#we can use ordiplot to create a plot with labels

ordiplot(example_NMDS,type="n")
orditorp(example_NMDS,display="species",col="red",air=0.01)
orditorp(example_NMDS,display="sites",cex=1.25,air=0.01)


treat=c(rep("Treatment1",5),rep("Treatment2",5))
ordiplot(example_NMDS,type="n")
ordiellipse(example_NMDS,groups=treat,draw="polygon",col="grey90",label=F)
orditorp(example_NMDS,display="species",col="red",air=0.01)
orditorp(example_NMDS,display="sites",col=c(rep("green",5),rep("blue",5)),
   air=0.01,cex=1.25)
```
